Resolving the measurement uncertainty paradox in ecological management

### 2018-12-28 manuscripts/manuscript.Rmd
rebuild manuscript

---
title: "Resolving the measurement uncertainty paradox in ecological management"

author:
  - name: "Authors removed during peer review"

#  - name: "Milad Memarzadeh"
#    affiliation: a
#  - name: "Carl Boettiger"
#    affiliation: a, 1
#address:
#  - code: a
#    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"

abstract: |
  Ecological management and decision-making typically focus on uncertainty about the future, but surprisingly little is known about how to account for uncertainty of the present: that is, the realities of having only partial or imperfect measurements.  Our primary paradigms for handling decisions under uncertainty -- the precautionary principle and optimal control -- have so far given contradictory results.  This paradox is best illustrated in the example of fisheries management, where many ideas that guide thinking about ecological decision making were first developed.  We find that simplistic optimal control approaches have repeatedly concluded that a manager should increase catch quotas when faced with greater uncertainty about the fish biomass. Current best practices take a more precautionary approach, decreasing catch quotas by a fixed amount to account for uncertainty.  Using comparisons to both simulated and historical catch data, we find that neither approach is sufficient to avoid stock collapses under moderate observational uncertainty.  Using partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value.  These results have significant implications for how we handle uncertainty in both fisheries and ecological management more generally.

keywords:
  - POMDP,
  - measurement uncertainty,
  - decision theory,
  - fisheries,
  - conservation

lineno: true

bibliography: citations.bib
#journal: "The American Naturalist"
date: "`r Sys.Date()`"
csl: the-american-naturalist.csl
output: rticles::elsevier_article
layout: 3p # review # review = doublespace, 3p = singlespace, 5p = two-column
header-includes: |
        \journal{The American Naturalist}
        \usepackage{lineno}
        \linenumbers
        \modulolinenumbers[3]
        \usepackage{setspace}
        \doublespacing

@@

- Transition process (state equation): $T(x_t, x_{t+1}, a_t)$: the probability that a system is in state $x_{t+1}$ at time $t+1$ given that it began in state $x_t$ at time $t$ and the manager took action $a_t$.  In our context, this relationship is given by the Gordon-Schaefer stock recruitment function $f$ with normally distributed growth uncertainty $x_{t+1} \sim \mathcal{N}(f(x_t,a_t), \sigma_g)$, truncated at zero to exclude negative population sizes. The action $a_t$ represents the harvest quota set for the fishery. Attempting to implement a quota the exceeds the true population size results in collapse of the fishery.  As @Clark1986 eloquently argues, assuming that extinction is impossible regardless of fishing intensity would unreasonably bias the decision problem.

- Observation function: $O(x_t,y_t,a_{t-1})$ the probability of observing state $y_t$ given a system in state $x_t$.  In principle, the action chosen can influence the precision of the observation. In our case, we simply assume normally distributed errors around the true state, $y_t \sim \mathcal{N}(x_t, \sigma_m)$, truncated at zero to exclude negative population sizes.

- Utility function: $U(x_t,a_t)$, the utility received at time $t$ for taking action $a_t$, given that the system is in state $x_t$.  For simplicity of analysis, we will simply set the utility to be equal to the harvested stock: $U(x_t, a_t) = \min(x_t, a_t)$, indicating that realized harvest cannot be negative.  This choice ensures that in the case of no uncertainty ($\sigma_g = \sigma_m = 0$), the optimal solution matches that expected under a simple MSY calculation. More realistic utility functions may include diminishing returns with increasing harvest (supply and demand effects), and the cost of fishing, both of which act to suppress large harvests. By focusing on a simple utility we can be sure that our comparison to MSY is driven by the treatment of uncertainty rather than merely differing economic assumptions.

@@

Is this approach too complex to be feasible? The importance of simple and effective rules of thumb in conservation management has been well documented [e.g. @Chades2011]. Sophisticated and computationally intensive approaches such as POMDP may seem improbable at a time when many areas of natural resource management, even simple, rule-of-thumb based methods struggle to take root.  There is no doubt that the solution method needed here is computationally intensive even for the simple model considered here, and will only become more so under greater ecological complexity.  Indeed, it was precisely such computational limitations that led to the shortcuts in work such as @Clark1986, with the ensuing potentially misleading conclusions. While in the following decades ecologists largely turned away from computationally intensive optimization required for even simple models with stochastic dynamic programming, (while continuing to construct ever more complicated hierarchical models), engineers have chipped away at the decision problem to reveal algorithms such as SARSOP that make much larger dynamic programming problems tractable.  Yes, we may need even more sophisticated algorithms to accommodate greater ecological complexity, yet we believe such numerical obstacles are no excuse for management by half measures.  Rather, our example illustrates how ecologists and managers can learn from and leverage the tools developed in other disciplines to better face the challenges of management in a complex and changing world.

>>>>

---
title: "Resolving the measurement uncertainty paradox in ecological management"
author:
  - name: "Milad Memarzadeh"
    affiliation: a
  - name: "Carl Boettiger"
    affiliation: a, 1
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  Ecological management and decision-making typically focus on uncertainty about the future, but surprisingly little is known about how to account for uncertainty of the present: that is, the realities of having only partial or imperfect measurements.  Our primary paradigms for handling decisions under uncertainty -- the precautionary principle and optimal control -- have so far given contradictory results.  This paradox is best illustrated in the example of fisheries management, where many ideas that guide thinking about ecological decision making were first developed.  We find that simplistic optimal control approaches have repeatedly concluded that a manager should increase catch quotas when faced with greater uncertainty about the fish biomass. Current best practices take a more precautionary approach, decreasing catch quotas by a fixed amount to account for uncertainty.  Using comparisons to both simulated and historical catch data, we find that neither approach is sufficient to avoid stock collapses under moderate observational uncertainty.  Using partially observed Markov decision process (POMDP) methods, we demonstrate how this paradox arises from flaws in the standard theory, which contributes to over-exploitation of fisheries and increased probability of economic and ecological collapse.  In contrast, we find POMDP-based management avoids such over-exploitation while also generating higher economic value.  These results have significant implications for how we handle uncertainty in both fisheries and ecological management more generally.
keywords:
  - POMDP,
  - measurement uncertainty,
  - decision theory,
  - fisheries,
  - conservation
bibliography: citations.bib
date: "`r Sys.Date()`"
csl: the-american-naturalist.csl
output: rticles::elsevier_article
layout: 3p # review # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \journal{The American Naturalist}
  \usepackage[nomarkers]{endfloat}

@@

**Transition** process (state equation): $T(x_t, x_{t+1}, a_t)$: the probability that a system is in state $x_{t+1}$ at time $t+1$ given that it began in state $x_t$ at time $t$ and the manager took action $a_t$.  In our context, this relationship is given by the Gordon-Schaefer stock recruitment function $f$ with normally distributed growth uncertainty $x_{t+1} \sim \mathcal{N}(f(x_t,a_t), \sigma_g)$, truncated at zero to exclude negative population sizes. The action $a_t$ represents the harvest quota set for the fishery. Attempting to implement a quota the exceeds the true population size results in collapse of the fishery.  As @Clark1986 eloquently argues, assuming that extinction is impossible regardless of fishing intensity would unreasonably bias the decision problem.

**Observation** function: $O(x_t,y_t,a_{t-1})$ the probability of observing state $y_t$ given a system in state $x_t$.  In principle, the action chosen can influence the precision of the observation. In our case, we simply assume normally distributed errors around the true state, $y_t \sim \mathcal{N}(x_t, \sigma_m)$, truncated at zero to exclude negative population sizes.

**Utility** function: $U(x_t,a_t)$, the utility received at time $t$ for taking action $a_t$, given that the system is in state $x_t$.  For simplicity of analysis, we will simply set the utility to be equal to the harvested stock: $U(x_t, a_t) = \min(x_t, a_t)$, indicating that realized harvest cannot be negative.  This choice ensures that in the case of no uncertainty ($\sigma_g = \sigma_m = 0$), the optimal solution matches that expected under a simple MSY calculation. More realistic utility functions may include diminishing returns with increasing harvest (supply and demand effects), and the cost of fishing, both of which act to suppress large harvests. By focusing on a simple utility we can be sure that our comparison to MSY is driven by the treatment of uncertainty rather than merely differing economic assumptions.

@@

Is this approach too complex to be feasible? The importance of simple and effective rules of thumb in conservation management has been well documented [e.g. @Chades2011]. Sophisticated and computationally intensive approaches such as POMDP may seem improbable at a time when many areas of natural resource management, even simple, rule-of-thumb based methods struggle to take root.  There is no doubt that the solution method needed here is computationally intensive even for the simple model considered here, and will only become more so under greater ecological complexity.  Indeed, it was precisely such computational limitations that led to the shortcuts in work such as @Clark1986, with the ensuing potentially misleading conclusions. While in the following decades ecologists largely turned away from computationally intensive optimization required for even simple models with stochastic dynamic programming, (while continuing to construct ever more complicated hierarchical models), engineers have chipped away at the decision problem to reveal algorithms such as SARSOP that make much larger dynamic programming problems tractable.  We may need even more sophisticated algorithms to accommodate greater ecological complexity, yet we believe such numerical obstacles are no excuse for management by half measures.  Rather, our example illustrates how ecologists and managers can learn from and leverage the tools developed in other disciplines to better face the challenges of management in a complex and changing world.

### 2018-12-28 manuscripts/manuscript.tex
rebuild manuscript

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header
\journal{The American Naturalist}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[3]
\usepackage{setspace}
\doublespacing

@@

  \title{Resolving the measurement uncertainty paradox in ecological management}
    \author[]{Authors removed during peer review}

@@

\begin{itemize}
\item
  Transition process (state equation): \(T(x_t, x_{t+1}, a_t)\): the
  probability that a system is in state \(x_{t+1}\) at time \(t+1\)
  given that it began in state \(x_t\) at time \(t\) and the manager
  took action \(a_t\). In our context, this relationship is given by the
  Gordon-Schaefer stock recruitment function \(f\) with normally
  distributed growth uncertainty
  \(x_{t+1} \sim \mathcal{N}(f(x_t,a_t), \sigma_g)\), truncated at zero
  to exclude negative population sizes. The action \(a_t\) represents
  the harvest quota set for the fishery. Attempting to implement a quota
  the exceeds the true population size results in collapse of the
  fishery. As Clark and Kirkwood (1986) eloquently argues, assuming that
  extinction is impossible regardless of fishing intensity would
  unreasonably bias the decision problem.
\item
  Observation function: \(O(x_t,y_t,a_{t-1})\) the probability of
  observing state \(y_t\) given a system in state \(x_t\). In principle,
  the action chosen can influence the precision of the observation. In
  our case, we simply assume normally distributed errors around the true
  state, \(y_t \sim \mathcal{N}(x_t, \sigma_m)\), truncated at zero to
  exclude negative population sizes.
\item
  Utility function: \(U(x_t,a_t)\), the utility received at time \(t\)
  for taking action \(a_t\), given that the system is in state \(x_t\).
  For simplicity of analysis, we will simply set the utility to be equal
  to the harvested stock: \(U(x_t, a_t) = \min(x_t, a_t)\), indicating
  that realized harvest cannot be negative. This choice ensures that in
  the case of no uncertainty (\(\sigma_g = \sigma_m = 0\)), the optimal
  solution matches that expected under a simple MSY calculation. More
  realistic utility functions may include diminishing returns with
  increasing harvest (supply and demand effects), and the cost of
  fishing, both of which act to suppress large harvests. By focusing on
  a simple utility we can be sure that our comparison to MSY is driven
  by the treatment of uncertainty rather than merely differing economic
  assumptions.
\end{itemize}

@@

Is this approach too complex to be feasible? The importance of simple
and effective rules of thumb in conservation management has been well
documented (e.g. Chadès et al. 2011). Sophisticated and computationally
intensive approaches such as POMDP may seem improbable at a time when
many areas of natural resource management, even simple, rule-of-thumb
based methods struggle to take root. There is no doubt that the solution
method needed here is computationally intensive even for the simple
model considered here, and will only become more so under greater
ecological complexity. Indeed, it was precisely such computational
limitations that led to the shortcuts in work such as Clark and Kirkwood
(1986), with the ensuing potentially misleading conclusions. While in
the following decades ecologists largely turned away from
computationally intensive optimization required for even simple models
with stochastic dynamic programming, (while continuing to construct ever
more complicated hierarchical models), engineers have chipped away at
the decision problem to reveal algorithms such as SARSOP that make much
larger dynamic programming problems tractable. Yes, we may need even
more sophisticated algorithms to accommodate greater ecological
complexity, yet we believe such numerical obstacles are no excuse for
management by half measures. Rather, our example illustrates how
ecologists and managers can learn from and leverage the tools developed
in other disciplines to better face the challenges of management in a
complex and changing world.

>>>>

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

\journal{The American Naturalist}
\usepackage[nomarkers]{endfloat}

@@

  \title{Resolving the measurement uncertainty paradox in ecological management}
    \author[a]{Milad Memarzadeh}

    \author[a, 1]{Carl Boettiger}

      \address[a]{Dept of Environmental Science, Policy, and Management, University of
California Berkeley, Berkeley CA 94720-3114, USA}

@@

\textbf{Transition} process (state equation): \(T(x_t, x_{t+1}, a_t)\):
the probability that a system is in state \(x_{t+1}\) at time \(t+1\)
given that it began in state \(x_t\) at time \(t\) and the manager took
action \(a_t\). In our context, this relationship is given by the
Gordon-Schaefer stock recruitment function \(f\) with normally
distributed growth uncertainty
\(x_{t+1} \sim \mathcal{N}(f(x_t,a_t), \sigma_g)\), truncated at zero to
exclude negative population sizes. The action \(a_t\) represents the
harvest quota set for the fishery. Attempting to implement a quota the
exceeds the true population size results in collapse of the fishery. As
Clark and Kirkwood (1986) eloquently argues, assuming that extinction is
impossible regardless of fishing intensity would unreasonably bias the
decision problem.

\textbf{Observation} function: \(O(x_t,y_t,a_{t-1})\) the probability of
observing state \(y_t\) given a system in state \(x_t\). In principle,
the action chosen can influence the precision of the observation. In our
case, we simply assume normally distributed errors around the true
state, \(y_t \sim \mathcal{N}(x_t, \sigma_m)\), truncated at zero to
exclude negative population sizes.

\textbf{Utility} function: \(U(x_t,a_t)\), the utility received at time
\(t\) for taking action \(a_t\), given that the system is in state
\(x_t\). For simplicity of analysis, we will simply set the utility to
be equal to the harvested stock: \(U(x_t, a_t) = \min(x_t, a_t)\),
indicating that realized harvest cannot be negative. This choice ensures
that in the case of no uncertainty (\(\sigma_g = \sigma_m = 0\)), the
optimal solution matches that expected under a simple MSY calculation.
More realistic utility functions may include diminishing returns with
increasing harvest (supply and demand effects), and the cost of fishing,
both of which act to suppress large harvests. By focusing on a simple
utility we can be sure that our comparison to MSY is driven by the
treatment of uncertainty rather than merely differing economic
assumptions.

@@

Is this approach too complex to be feasible? The importance of simple
and effective rules of thumb in conservation management has been well
documented (e.g. Chadès et al. 2011). Sophisticated and computationally
intensive approaches such as POMDP may seem improbable at a time when
many areas of natural resource management, even simple, rule-of-thumb
based methods struggle to take root. There is no doubt that the solution
method needed here is computationally intensive even for the simple
model considered here, and will only become more so under greater
ecological complexity. Indeed, it was precisely such computational
limitations that led to the shortcuts in work such as Clark and Kirkwood
(1986), with the ensuing potentially misleading conclusions. While in
the following decades ecologists largely turned away from
computationally intensive optimization required for even simple models
with stochastic dynamic programming, (while continuing to construct ever
more complicated hierarchical models), engineers have chipped away at
the decision problem to reveal algorithms such as SARSOP that make much
larger dynamic programming problems tractable. We may need even more
sophisticated algorithms to accommodate greater ecological complexity,
yet we believe such numerical obstacles are no excuse for management by
half measures. Rather, our example illustrates how ecologists and
managers can learn from and leverage the tools developed in other
disciplines to better face the challenges of management in a complex and
changing world.
